---
title: "Exploring The Structure and Dependencies of An R Package"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Exploring The Structure and Dependencies of An R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupVignette, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.align = 'center'
)
```


<!--  STYLE SET UP FOR TWO COLUMN SECTIONS  -->
<style>
.row:after {
    content: "";
    display: table;
    clear: both;
}
.column-left{
  float: left;
  width: 48%;
  text-align: justify;
  text-justify: inter-word;
  margin-right: 4%;
}
.column-right{
  float: right;
  width: 48%;
  text-align: justify;
  text-justify: inter-word;
}
</style>


<!--  CODE BELOW HERE WILL RENDER  -->

`pkgnet` is an R library designed for the analysis of R libraries! The goal of the package is to build a graph representation of a package and its dependencies to inform a variety of activities, including:

- prioritizing functions to unit test based on their centrality
- examining the recursive dependencies you are taking on by using a given package
- exploring the structure of a new package provided by a coworker or downloaded from the internet

Below is a brief tour of `pkgnet` and its features. 

***
# Dependencies As a Graph
`pkgnet` represents both package dependencies and function dependencies as directed graphs. Before we look at the output of `pkgnet`, here are few core concepts to keep in mind.

```{r whatIsDepGraph, echo=FALSE, message=FALSE, results='markup', fig.height=3, fig.width=3, fig.cap = "Example Dependency Graph"}
nodes <- data.frame(
  id = 1:4
  , label = LETTERS[1:4]
  , x = c(1,3,2,4)
  , y = c(1,1,2,3)
  , level = c(1,1,2,3)
  )

edges <- data.frame(
  from = c(1,2,3,2)
  , to = c(3,3,4,4)
  )

g <- visNetwork::visNetwork(nodes = nodes
                       , edges = edges
                       , width = "100%"
                       )
g <- visNetwork::visNodes(graph = g
                          , shape = "circle"
                          , font = list(size = 25
                                        , bold = TRUE
                                        , align = 'center'
                                        )
                       )
g <- visNetwork::visEdges(graph = g
                          , arrows = "to"
                       )
g <- visNetwork::visHierarchicalLayout(graph = g
                      , direction = "DU"
                      )
g <- visNetwork::visInteraction(graph = g
                                , dragNodes = TRUE
                                , dragView = TRUE
                                , zoomView = FALSE)
g
```

## Dependency
Functions or Dependencies are represented as nodes, and their dependent relationships are represented as edges (a.k.a. arcs or arrows).  The direction of the edge points towards the node that is dependent on the other node.  

In the example dependency graph above:

* **C** is dependent upon both **A** and **C**.
* **D** is dependent upon both **C** and **B**.
* **D** is indirectly dependent upon **A** through **C** via the transitive property.

## Descendants
The descendants of a node are all subsequent nodes that depend on that node, either directly or via the transitive property.

In the example dependency graph above:

* **C** and **D** are descendants of **A**.
* **B** is **NOT** a descendant of **A**.

***
# Running pkgnet

`pkgnet` can analyze any R package locally installed.  Run `installed.packages()` to see the full list of packages installed on your system.  For this example, let's say we are analyzing a custom built package, `baseballstats`.

```{r makeDemoPackage, include=FALSE}
devtools::install_local(path = system.file('baseballstats',package="pkgnet")
                        , force=TRUE
                        , lib = .libPaths()
                        )
```

To analyze `baseballstats`, run the following two lines of code: 

```{r pkgnetRunFirst, message=FALSE, warning=FALSE, results='hide'}
library(pkgnet)
report1 <- CreatePackageReport(pkg_name = "baseballstats")
```

That's it! You have generated a lot of valuable information with that one call for an installed package.  

However, if the full source repository for the package is available on your system, you can supplement this report with other information such as code coverage from [covr](https://CRAN.R-project.org/package=covr).  To do so, specify the path to the repository in `CreatePackageReport`.  


```{r pkgnetRunFAKE, eval=FALSE}
library(pkgnet)
report2 <- CreatePackageReport(
  pkg_name = "baseballstats"
  , pkg_path = <path to the repo>
)
```

```{r pkgnetRunSecond, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
library(pkgnet)
report2 <- CreatePackageReport(
  pkg_name = "baseballstats"
  , pkg_path = system.file('baseballstats',package="pkgnet")
)
```

***
# Examining The Results

An HTML report has been created with the pertinent information and a list object is available with the same information and more.  

## The Report

An HTML report has been created, and its location is specified in the messages in the terminal.  

This report has three sections: 

1. **Package Summary** -- general information about the package and package level statistics   
2. **Dependency Network** -- information regarding the packages upon which the current package under analysis relies upon    
3. **Function Network** -- information regarding the functions within the current package under analysis and their inter dependencies     

Each section has helpful tables and visuals.  

As a sample, here's how the **Function Network Visualization** looks for `baseballstats`: 

<!--  SIDE BY SIDE START  -->
<div class="row"> <!--  Div to Wrap columns -->

<div class="column-left">
#### Default
```{r demoVis1, echo=FALSE, fig.width=3, fig.height=3}
report1$FunctionReporter$graph_viz
```
All Functions and their dependencies are visible.  For example, we can see that both `batting_avg` and `slugging_avg` functions depend upon the `at_bats` function.  

We also see that nothing depends on the `on_base_pct` function.  This might be valuable information to an R package developer. 
</div> <!--  End column-left -->

<div class="column-right">
#### With Coverage Information 
```{r demoVis2, echo=FALSE, , fig.width=3, fig.height=3}
report2$FunctionReporter$graph_viz
```
Same as the default visualization except we can see coverage information as well (Red = 0%, Green = 100%). 

It appears the function with the most dependencies, `at_bats`, is well covered.  However, no other functions are covered by unit tests.
</div> <!--  End column-right  -->

</div> <!--  End Div to Wrap columns -->
<!--  SIDE BY SIDE END -->

**Check out the full HTML report for more results**  

## The List Object

The `CreatePackageReport()` function returns a list with three items: 

1. SummaryReporter  
2. DependencyReporter  
3. FunctionReporter  

Each items contains information visible in the report *and more*.  We can use this information for a more detailed analysis of the results and/or more easily incorporate `pkgnet` results into other R processes.

Here are a few notable items available within the list object.

### Node Information
Both the `DependencyReporter` and the `FunctionReporter` contain metrics about their package dependencies or functions (a.k.a network nodes) in a `nodes` table.  

```{r nodes}
dim(report2$FunctionReporter$nodes)
names(report2$FunctionReporter$nodes)
```

Note, a few of these metrics provided by default are from the field of [Network Theory](https://en.wikipedia.org/wiki/Network_theory).  You can leverage the **Network Object** described below to derive many more.     

### Network Measures 
Both the `DependencyReporter` and the `FunctionReporter` contain measures based on their network structure in a `network_measures` list. 

```{r networkMeasures}
report2$FunctionReporter$network_measures
```

### Network Object
Both the `DependencyReporter` and the `FunctionReporter` are available as [igraph](http://igraph.org/r/) objects named `pkg_graph`

```{r networkObj}
report2$FunctionReporter$pkg_graph
```

***
# A Deeper Look
With the reports and objects produced by `pkgnet` by default, there is plenty to inform us on the inner workings of an R package.  However, we may want to know MORE! Since the [igraph](http://igraph.org/r/) objects are available, we can leverage those graphs for further analysis.  
     
In this section, let's examine a larger R package, such as [lubridate](http://lubridate.tidyverse.org/).
       
If you would like to follow along with the examples in this section, run these commands in your terminal to download and install `lubridate`^[Examples from version 1.7.3 of Lubridate].  
```{r bashText, engine='bash', eval=FALSE}
# Create a temporary workspace
mkdir -p ~/pkgnet_example && cd ~/pkgnet_example

# Grab the lubridate source code
git clone https://github.com/tidyverse/lubridate
cd lubridate

# If you want the examples to match exactly
git reset --hard 9797d69abe1574dd89310c834e52d358137669b8

# Install it
Rscript -e "devtools::install()"
```

## Coverage of Most Referenced Functions
Let's examine the relationship between a function's total number of descendants and the unit test coverage of that function's code.

<!--  Faked Since lubridate not guaranteed to be installed at time of report creation -->
```{r fakeDetail1, eval=FALSE}
# Run pkgnet
library(pkgnet)
report2 <- CreatePackageReport(
    pkg_name = "lubridate"
    , pkg_path = "~/pkgnet_example/lubridate"
)

# Extract Nodes Table
funcNodes <- report2$FunctionReporter$nodes

# List Coverage For Most Referenced Functions
mostRef <- funcNodes[order(numDescendants, decreasing = TRUE)][1:10]
mostRef[,list(`Function` = node
              , `Descendant Count` = numDescendants
              , `Coverage Ratio` = coverageRatio
              , `Total Lines` = totalLines)]

```

|Function                | Descendant Count| Coverage Ratio| Total Lines|
|:-----------------------|----------------:|--------------:|-----------:|
|divide_period_by_period |               39|              1|           2|
|days                    |               22|              1|           1|
|check_duration          |               15|              0|           1|
|as.POSIXt               |               13|              0|           1|
|eweeks                  |               13|              0|           2|
|check_interval          |               12|              0|          11|
|date<-                  |               12|             NA|          NA|
|add_months              |               10|              1|           4|
|ceil_multi_unit         |               10|              1|           1|
|am                      |                6|              1|           1|

Inspecting results such as these can help an R package developer decide which function to cover with unit tests next.  

In this case, `check_duration`, one of the most referenced functions (either directly or indirectly), is not covered by unit tests.  However, it appears to be a simple one line function that may not be necessary to cover in unit testing.  `check_interval`, on the other hand, might benefit from some unit test coverage as it is a larger, uncovered function with a similar number of dependencies.

## Discovering Similar Functions
Looking at that same large package, let's say we wanted to explore options for consolidating functions.  One approach might be to explore consolidating functions that share the same dependencies.  In that case, we could use the `igraph` object to highlight functions with the same dependencies via [Jaccard similarity](https://en.wikipedia.org/wiki/Jaccard_index). 

```{r fakeDetail2, eval=FALSE}
# Get igraph object
funcGraph <- report2$FunctionReporter$pkg_graph
funcNames <- igraph::vertex_attr(funcGraph, name = "name")

# Jaccard Similarity
sim <- igraph::similarity(graph = funcGraph
                          , mode = "in"
                          , method = "jaccard")
diag(sim) <- 0
sim[sim < 1] <- 0

simGraph <- igraph::graph_from_adjacency_matrix(adjmatrix = sim, mode = "undirected")

# Find groups with same dependencies (similarity == 1)
sameDeps <- igraph::max_cliques(graph = simGraph
                                , min = 2
                                )

# Write results
for (i in seq_along(sameDeps)) {
         cat(paste0("Group ", i, ": "))
         cat(paste(funcNames[as.numeric(sameDeps[[i]])], collapse = ", "))
         cat("\n")
}
```
  
```{r resultFromFake, echo=FALSE, results='markup'}
cat("Group 1: stamp_time, stamp_date
Group 2: ms, hm
Group 3: new_interval, %--%, int_diff
Group 4: floor_date, quarter, semester
Group 5: picoseconds, microseconds, nanoseconds, milliseconds
Group 6: weeks, days, years, seconds_to_period, seconds, new_period, minutes, hours
Group 7: yq, dmy, ymd_hms, ymd_hm, ymd_h, ymd, ydm_hms, ydm_hm, ydm_h, ydm, pretty_dates, parse_date_time2, parse_date_time, myd, mdy_hms, mdy_hm, mdy_h, mdy, local_time, fast_strptime, dym, dmy_hms, dmy_hm, dmy_h
"
)
```

Now, we have identified seven different groups of functions within [lubridate](http://lubridate.tidyverse.org/) that share the _exact same_ dependencies.  We could explore each group of functions for potential consolidation. 



```{r removeDemoPackage, include=FALSE}
devtools::uninstall(system.file('baseballstats',package="pkgnet"))
```

