---
title: "Exploring the Structure and Dependencies of an R Package"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Exploring the Structure and Dependencies of an R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupVignette, include = FALSE}

## NOTE:    Vignettes are built within their own environment. 
##          Therefore, changing the library paths (libpaths) here will not change the global libpaths.

# Get test library path
testLibPath <- tempdir()

# Get current library paths
origLibPaths <- .libPaths()

# Create new library paths for TESTING
.libPaths(new = c(testLibPath, origLibPaths))

# Create packages within new library
pkgnet:::.BuildTestLib(targetLibPath = testLibPath)

knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.align = 'center',
  out.width='100%'
)
```


<!--  STYLE SET UP FOR TWO COLUMN SECTIONS  -->
<style>
.row:after {
    content: "";
    display: table;
    clear: both;
}
.column-left{
  float: left;
  width: 48%;
  text-align: justify;
  text-justify: inter-word;
  margin-right: 4%;
}
.column-right{
  float: right;
  width: 48%;
  text-align: justify;
  text-justify: inter-word;
}
</style>


<!--  CODE BELOW HERE WILL RENDER  -->

`pkgnet` is an R package designed for the analysis of R packages! The goal of the package is to build graph representations of a package's various types of dependencies. This can inform a variety of activities, including:

- prioritizing functions to unit test based on their centrality or influence
- examining the recursive dependencies you are taking on by using a given package
- exploring the structure of a new package provided by a coworker or downloaded from the internet

Below is a brief tour of `pkgnet` and its features. 

***
# Packages as a Graph
`pkgnet` represents aspects of R packages as graphs. The two default reporters, which we will discuss in this vignette, model their respective aspects as directed graphs: a package's dependencies on other packages, and the interdependencies of functions within a package. Before we look at the output of `pkgnet`, here are few core concepts to keep in mind.

```{r whatIsDepGraph, echo=FALSE, message=FALSE, fig.height=3, results='markup', fig.cap = "Example Dependency Graph"}
nodes <- data.frame(
  id = 1:4
  , label = LETTERS[1:4]
  , x = c(1,3,2,4)
  , y = c(1,1,2,3)
  , level = c(1,1,2,3)
  )

edges <- data.frame(
  from = c(3,3,4,4)
  , to = c(1,2,3,2)
  )

g <- visNetwork::visNetwork(nodes = nodes
                       , edges = edges
                       , width = "100%"
                       )
g <- visNetwork::visNodes(graph = g
                          , shape = "circle"
                          , font = list(size = 25
                                        , bold = TRUE
                                        , align = 'center'
                                        )
                       )
g <- visNetwork::visEdges(graph = g
                          , arrows = "to"
                       )
g <- visNetwork::visHierarchicalLayout(graph = g
                      , direction = "DU"
                      , sortMethod = "directed"
                      )
g <- visNetwork::visInteraction(graph = g
                                , dragNodes = TRUE
                                , dragView = TRUE
                                , zoomView = FALSE)
g
```

## Dependency

Units of the analysis are represented as **nodes**, and their dependency relationships are represented as **edges** (a.k.a. arcs or arrows). In `pkgnet`, the nodes could be functions in the package you are examining, or other packages that the package depends on. The direction of edges point in the direction of dependency---the tail node depends on the head node.^[Edge direction was previously Independent -> Dependent. It was changed to Dependent -> Independent in version v0.3.0. The new convention follows the [Unified Modeling Language (UML)](https://en.wikipedia.org/wiki/Dependency_\(UML\)) framework, a widely used standard for software system modeling.] 

In the example dependency graph above:

* **C** depends on both **A** and **B**.
* **D** depends both **C** and **B**.
* **D** indirectly depends upon **A** through **C** via the transitive property.
* **B** does not depend on **A** as there is no path from **A** to **B** by following the edges.

Following the direction of the edges allows you to figure out the **dependencies** of a node---the other nodes that it depends on. On the flip side, tracing the edges backwards allows you to figure out the **reverse dependencies** (i.e., dependents) of a node---the other nodes that depend on it.

***
# Running pkgnet

`pkgnet` can analyze any R package locally installed.  (Run `installed.packages()` to see the full list of packages installed on your system.)  For this example, let's say we are analyzing a custom built package, `baseballstats`.

To analyze `baseballstats`, run the following two lines of code: 

```{r pkgnetRunFirst, eval=FALSE}
library(pkgnet)
report1 <- CreatePackageReport(pkg_name = "baseballstats")
```

That's it! You have generated a lot of valuable information with that one call for an installed package.  

However, if the full source repository for the package is available on your system, you can supplement this report with other information such as code coverage from [covr](https://CRAN.R-project.org/package=covr).  To do so, specify the path to the repository in `CreatePackageReport`.  


```{r pkgnetRunFAKE, eval=FALSE}
library(pkgnet)
report2 <- CreatePackageReport(
  pkg_name = "baseballstats"
  , pkg_path = <path to the repo>
)
```

***
# Examining the Results

`CreatePackageReport` has written an HTML report with the pertinent information, and it also returned a list object with the same information and more.  

## The Report

The location of the HTML report is specified in the messages in the terminal. 

This report has three sections: 

1. **Package Summary** -- general information about the package and package level statistics
2. **Dependency Network** -- information regarding the packages upon which the current package under analysis depends upon
3. **Function Network** -- information regarding the functions within the current package under analysis and their interdependencies

Each section has helpful tables and visuals. 

As a sample, here's how the **Function Network Visualization** looks for `baseballstats`: 

<!--  SIDE BY SIDE START  -->
<div class="row"> <!--  Div to Wrap columns -->

<div class="column-left">
#### Default
```{r demoVis1, fig.height=3, message=FALSE, warning=FALSE, echo=FALSE}
pkgnet:::silence_logger()
funcReporter1 <- pkgnet::FunctionReporter$new()
funcReporter1$set_package('baseballstats')
funcReporter1$layout_type <- "layout_as_tree"
g <- visNetwork::visHierarchicalLayout(
    graph = funcReporter1$graph_viz
    , direction = "UD"
    , sortMethod = "directed"
    , edgeMinimization = FALSE
)
g <- visNetwork::visInteraction(graph = g
                                , dragNodes = TRUE
                                , dragView = TRUE
                                , zoomView = FALSE)
g
```
All functions and their dependencies are visible.  For example, we can see that both `batting_avg` and `slugging_avg` functions depend upon the `at_bats` function.  

We also see that nothing depends on the `on_base_pct` function.  This might be valuable information to an R package developer. 
</div> <!--  End column-left -->

<div class="column-right">
#### With Coverage Information 
```{r demoVis2, fig.height=3, message=FALSE, warning=FALSE, echo=FALSE}
pkgnet:::silence_logger()
funcReporter2 <- pkgnet::FunctionReporter$new()
funcReporter2$layout_type <- "layout_as_tree"
funcReporter2$set_package(
    pkg_name = "baseballstats"
    , pkg_path = system.file('baseballstats',package="pkgnet")
)
funcReporter2$calculate_default_measures()
g <- visNetwork::visHierarchicalLayout(
    graph = funcReporter2$graph_viz
    , direction = "UD"
    , sortMethod = "directed"
    , edgeMinimization = FALSE
)
g <- visNetwork::visInteraction(graph = g
                                , dragNodes = TRUE
                                , dragView = TRUE
                                , zoomView = FALSE)
g
```
Same as the default visualization except we can see coverage information as well (Pink = 0%, Green = 100%). 

It appears the function with the most dependencies, `at_bats`, is well covered.  However, no other functions are covered by unit tests.
</div> <!--  End column-right  -->

</div> <!--  End Div to Wrap columns -->
<!--  SIDE BY SIDE END -->

**Check out the full HTML report for more results**  

## The List Object

The `CreatePackageReport()` function returns a list with three items: 

1. SummaryReporter  
2. DependencyReporter  
3. FunctionReporter  

Each items contains information visible in the report *and more*.  We can use this information for a more detailed analysis of the results and/or more easily incorporate `pkgnet` results into other R processes.

Here are a few notable items available within the list object:

### Node Information
Both the `DependencyReporter` and the `FunctionReporter` contain metrics about their package dependencies or functions (a.k.a network nodes) in a `nodes` table.  
```{r mockPackageReport, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
# We initialized just the reporters because we didn't want to actually generate the full html report. So we'll put funcReporter2 into a list to mock the interface for the example
report2 <- list(FunctionReporter = funcReporter2)
```

```{r nodes}
dim(report2$FunctionReporter$nodes)
names(report2$FunctionReporter$nodes)
```

Note, a few of these metrics provided by default are from the field of [Network Theory](https://en.wikipedia.org/wiki/Network_theory).  You can leverage the **Network Object** described below to derive many more.     

### Network Measures 
Both the `DependencyReporter` and the `FunctionReporter` contain graph-level measures based on their network structure in a `network_measures` list. 

```{r networkMeasures}
report2$FunctionReporter$network_measures
```

### Network Graph Model Object
Both the `DependencyReporter` and the `FunctionReporter` have an object called `pkg_graph` that contains the graph model of their respective networks. This object has methods to calculate additional node-level and graph-level measures. It is powered by [igraph](http://igraph.org/r/), and the igraph object itself is directly accessible with `pkg_graph$igraph`. 

```{r networkObjAddtlMeasures}
report2$FunctionReporter$pkg_graph$node_measures(c('hubScore', 'authorityScore'))
```

```{r networkObj}
report2$FunctionReporter$pkg_graph$igraph
```

***
# A Deeper Look

With the reports and objects produced by `pkgnet` by default, there is plenty to inform us on the inner workings of an R package.  However, we may want to know MORE! Since the [igraph](http://igraph.org/r/) objects are available, we can leverage those graphs for further analysis.  
     
In this section, let's examine a larger R package, such as [lubridate](http://lubridate.tidyverse.org/).
       
If you would like to follow along with the examples in this section, run these commands in your terminal to download and install `lubridate`^[Examples from version 1.7.3 of Lubridate].  
```{r bashText, engine='bash', eval=FALSE}
# Create a temporary workspace
mkdir -p ~/pkgnet_example && cd ~/pkgnet_example

# Grab the lubridate source code
git clone https://github.com/tidyverse/lubridate
cd lubridate

# If you want the examples to match exactly
git reset --hard 9797d69abe1574dd89310c834e52d358137669b8

# Install it
Rscript -e "devtools::install()"
```

## Coverage of Most Depended-on Functions
Let's examine `lubridate`'s functions through the lens of each function's total number of dependents (i.e., the other functions that depend on it) and its code's unit test coverage. In our graph model for the `FunctionReporter`, the subgraph of paths leading into a given node is the set of functions that directly or indirectly depend on the function that node represents.  

<!--  Faked Since lubridate not guaranteed to be installed at time of report creation -->
```{r fakeDetail1, eval=FALSE}
# Run pkgnet
library(pkgnet)
report2 <- CreatePackageReport(
    pkg_name = "lubridate"
    , pkg_path = "~/pkgnet_example/lubridate"
)

# Extract Nodes Table
funcNodes <- report2$FunctionReporter$nodes

# List Coverage For Most Depended-on Functions
mostRef <- funcNodes[order(numRecursiveRevDeps, decreasing = TRUE),
                     .(node, numRecursiveRevDeps, coverageRatio, totalLines)
                     ][1:10]
```

```{r fakeDetail1Results, eval=FALSE}
#>             node numRecursiveRevDeps coverageRatio totalLines
#>  1:        month                  81             1          1
#>  2:           tz                  79             1          1
#>  3: reclass_date                  68             1          1
#>  4:         date                  67             1          1
#>  5:      is.Date                  60             1          1
#>  6:    is.POSIXt                  57             1          1
#>  7:         wday                  56             1          1
#>  8:   is.POSIXct                  55             1          1
#>  9:  .deprecated                  55             0         10
#> 10:      as_date                  52             1          1
```

Inspecting results such as these can help an R package developer decide which function to cover with unit tests next.  

In this case, `check_duration`, one of the most depended-on functions (either directly or indirectly), is not covered by unit tests.  However, it appears to be a simple one line function that may not be necessary to cover in unit testing.  `check_interval`, on the other hand, might benefit from some unit test coverage as it is a larger, uncovered function with a similar number of dependencies.

## Discovering Similar Functions
Looking at that same large package, let's say we want to explore options for consolidating functions.  One approach might be to explore consolidating functions that share the same dependencies.  In that case, we could use the `igraph` object to highlight functions with the same out-neighborhood via [Jaccard similarity](https://en.wikipedia.org/wiki/Jaccard_index). 

```{r fakeDetail2, eval=FALSE}
# Get igraph object
funcGraph <- report2$FunctionReporter$pkg_graph
funcNames <- igraph::vertex_attr(funcGraph, name = "name")

# Jaccard Similarity
sim <- igraph::similarity(graph = funcGraph
                          , mode = "out"
                          , method = "jaccard")
diag(sim) <- 0
sim[sim < 1] <- 0

simGraph <- igraph::graph_from_adjacency_matrix(adjmatrix = sim, mode = "undirected")

# Find groups with same out-neighbors (similarity == 1)
sameDeps <- igraph::max_cliques(graph = simGraph
                                , min = 2
                                )

# Write results
for (i in seq_along(sameDeps)) {
    cat(paste0("Group ", i, ": "))
    cat(paste(funcNames[as.numeric(sameDeps[[i]])], collapse = ", "))
    cat("\n")
}
```
  
```{r resultFromFake, echo=FALSE, results='markup'}
cat("Group 1: divisible_period, make_date
Group 2: parse_date_time2, fast_strptime
Group 3: .deprecated_fun, .deprecated_arg
Group 4: stamp_date, stamp_time
Group 5: epiweek, isoweek
Group 6: ms, hm
Group 7: quarter, semester
Group 8: am, .roll_hms
Group 9: modulo_interval_by_duration, modulo_interval_by_period
Group 10: .difftime_from_pieces, .duration_from_units
Group 11: divide_period_by_period, xtfrm.Period
Group 12: int_diff, %--%
Group 13: isoyear, epiyear
Group 14: nanoseconds, microseconds, picoseconds, milliseconds
Group 15: period_to_seconds, check_period, multiply_period_by_number, format.Period, divide_period_by_number, add_period_to_period
Group 16: myd, dmy, yq, ymd, dym, mdy, ydm
Group 17: hours, weeks, minutes, years, days, months.numeric, seconds, seconds_to_period
Group 18: C_force_tz, hour.default, mday.default, c.POSIXct, .mklt, yday.default, year.default, minute.default, second.default
Group 19: ehours, emilliseconds, eyears, eseconds, epicoseconds, enanoseconds, eminutes, olson_time_zones, edays, emicroseconds, eweeks
Group 20: dmy_h, ydm_hms, ymd_hms, dmy_hm, ymd_h, ydm_hm, ydm_h, dmy_hms, ymd_hm, mdy_hms, mdy_hm, mdy_h"
)
```

Now, we have identified twenty different groups of functions within [lubridate](http://lubridate.tidyverse.org/) that share the _exact same_ dependencies.  We could explore each group of functions for potential consolidation. 

```{r removeDemoPackage, include=FALSE}
utils::remove.packages(
    pkgs = c('baseballstats', 'sartre', 'pkgnet')
    , lib = testLibPath
)

# Just in case 
.libPaths(new = c(origLibPaths))
unlink(testLibPath)
```

